\chapter*{Acknowledgements}

People who know me more intimately know well that I have a
tendency for taking on many challenges at once. Over the
last three years, I (perhaps ambitiously, if not dubiously)
chose to concurrently pursue my Master's thesis, a Bachelor's
in Computer Science, and a Specialization in Artificial
Intelligence. This demanding series of endeavors proved to be
an invaluable learning experience, due in large part to the
steadfast support of my family, friends, and the wonderful
community of professors and students at the Institute of
Mathematics and Statistics of the University of São Paulo
(IME-USP), especially the LIAMF laboratory.

I would like to express my deepest gratitude to my advisor,
Professor Denis Deratani Mauá. Thank you not only for the
usual guidance and insightful support expected of a Master's
thesis, but also for continuously instigating my curiosity
about the world of Probabilistic Logic Programming and its
relation to Probabilistic Circuits. Your patience, expertise
in explaining complex concepts, sound advice, and overall
dedication to every piece of work related to yours, including
this thesis and the articles we co-authored, have been truly
foundational to my growth. I am especially grateful for your
constant availability and the stimulating discussions that
often ranged from early in the morning until late at night.

A special note of thanks goes to the members of both the
Qualifying and Defense boards. Your incisive feedback and
constructive discussions significantly helped me refine and
improve this work.

To my friends and colleagues at IME-USP, and especially within
the LIAMF lab, thank you for the support, the laughs, and all
the interesting moments we shared over the years. This journey
was made infinitely more enjoyable because of our community. A
special acknowledgement is reserved for Renato Lui Geh, who I
consider not only a dear friend but also a key mentor and
inspiration alongside Denis, capturing my attention in the fields
of Probabilistic Circuits, Neuro-Symbolic Artificial Intelligence,
and beyond.

Lastly, and most importantly, I thank my family for their unwavering
support and encouragement throughout this journey: my mother Ana,
my father André, and my brother Diogo. And, of course, a special and
warm thank you to both my grandmother Fátima and my grandfather Chico,
to whom this thesis is dedicated. Thank you all for everything.
