% Chapter 6

\chapter{Conclusion}

The proposed neuro-symbolic pipeline for argumentation mining leverages
probabilistic Answer Set Programming and knowledge compilation techniques to
enable efficient inference over complex argumentative structures. By grounding
PASP programs, applying Clark's completion, and compiling the resulting CNF into
sd-DNNF circuits, we aim to facilitate scalable reasoning in neuro-symbolic
learning loops, with potential applications in various domains requiring robust
argumentation analysis.

The proposed methods for efficiente inference using circuits are amenable to
integration with gradient-based learning frameworks, paving the way for future
work that combines neural networks with symbolic reasoning in a unified
architecture.

In our experimental evaluation \ref{subsec:empirical-analysis}, we observed that
\textsc{SharpSAT-TD} outperformed \textsc{C2D} and \textsc{D4} in
both compilation time and circuit size, when considering non-stratified PASP
instances, which are the focus of this work. This finding is aligned with our
expectations and previous studies in the literature \citep{EITER2024104109, kiesel2023knowledge}.
The superior performance of \textsc{SharpSAT-TD} can be attributed to its advanced top-down
compilation techniques, where \textsc{C2D} stands as a robust, but older
alternative, and \textsc{D4} shows very competitive performance in certain scenarios,
even surpassing \textsc{SharpSAT-TD} in stratified scenarios.

These results highlight the importance of selecting an appropriate knowledge
compiler for neuro-symbolic applications, specially given an specific domain,
such is the case of argumentation mining. The efficiency gains achieved by using
Knowledge Compilation can significantly enhance the practicality of
neuro-symbolic learning loops, enabling larger and more complex argumentation
graphs to be reasoned about effectively. Future work will focus on exploring
more classes of argumentations structures, specially those that take advantage
of the non-monotonic reasoning capabilities of PASP, and integrating the
compiled circuits into end-to-end neuro-symbolic learning frameworks, in order
to develop robust argumentation mining systems capable of handling real-world
data that current methods struggle with or cannot even model adequately.
